---
title: "Advanced topics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Token policy in depth

RSurvivEHR constructs the event vocabulary as:

```
Index  Token
─────  ─────────────────────────────────────────
  0    <PAD>   always present; padding_idx for the embedding layer
  1    <UNK>   when include_unk = TRUE (default)
  2    <CLS>   when include_cls_sep = TRUE
  3    <SEP>   when include_cls_sep = TRUE
  2+   clinical events, alphabetically sorted
```

`<PAD>` is always index 0 because SurvivEHR's `SplitDynamicEmbeddingLayer`
uses `padding_idx = 0` internally.

At prediction time any event code not present in `event_vocab` is replaced
by `<UNK>` (index 1), provided `include_unk = TRUE`.

---

## Age normalisation

All raw ages are divided by `time_scale` before entering the model:

$$a_{\text{norm}} = \frac{a_{\text{raw}}}{\texttt{time\_scale}}$$

The target time delta for fine-tuning is:

$$\delta_{\text{norm}} = \frac{a_{\text{target}} - a_{\text{last context}}}{\texttt{time\_scale}}$$

Use `time_scale = 1.0` (default) when ages are already in **years**.  
Use `time_scale = 1825.0` when ages are in **days since birth**, matching
FastEHR's `DAYS_SINCE_BIRTH` convention (1825 days ≈ 5 years).

The `time_scale` is stored in the model bundle and automatically applied at
fine-tune and predict time, so you only need to set it once in
`survivehr_config()`.

---

## Static covariates encoding

Static features are encoded into a flat numeric vector per patient:

| Column type | Encoding |
|---|---|
| Character / factor / < 80 % numeric | One-hot via `pd.get_dummies` |
| Fully numeric (int or float) | Passed through as-is |
| Patient missing from `static_df` | All zeros |

> **Column naming rule**: `patient_id` (or `PATIENT_ID`) is the only reserved
> name in the static table.  All other column names are **user-defined** —
> call them `SEX`, `imd`, `YearOfBirth`, or anything else.  The column names
> become the basis for encoded feature names (e.g. `SEX` → `SEX_F`, `SEX_M`).
> This is in contrast to the **events** and **targets** tables, which require
> fixed column names (`patient_id`, `event`, `age`, `value`) or their FastEHR
> UPPER_CASE aliases.  The only constraint for static columns is **consistency**:
> the exact same names must be used at pretrain, fine-tune, and predict time.

```{r}
static_covariates <- data.frame(
  patient_id     = 1:3,
  SEX            = c("M",     "F",     "M"),       # → SEX_F, SEX_M
  ETHNICITY      = c("White", "Asian", "White"),   # → ETHNICITY_Asian, ETHNICITY_White
  SMOKING_STATUS = c("Never", "Ex",    "Current"), # → SMOKING_STATUS_Current, ..., Never
  IMD            = c(3L, 1L, 5L),                 # numeric, passed through
  YEAR_OF_BIRTH  = c(1965L, 1972L, 1955L)         # numeric, passed through
)
```

The resulting static embedding for patient 1 (`M`, `White`, `Never`, IMD=3, YOB=1965)
would be:

```
[SEX_F=0, SEX_M=1, ETHNICITY_Asian=0, ETHNICITY_White=1,
 SMOKING_STATUS_Current=0, SMOKING_STATUS_Ex=0, SMOKING_STATUS_Never=1,
 IMD=3, YEAR_OF_BIRTH=1965]
```

The **same** encoding scheme (same column order, same dummy categories) is
derived at pre-train time and then **reused** at fine-tune and predict time
from the saved bundle.

---

## Using a fixed vocabulary

If you have pre-computed a vocabulary (e.g., from a previous run or from a
shared codelist), pass it as `event_vocab`:

```{r}
# Build vocab manually
my_vocab <- c("<PAD>" = 0L, "<UNK>" = 1L,
              "HYPERTENSION" = 2L, "METFORMIN" = 3L,
              "STATIN" = 4L, "T2D" = 5L)

pt_model <- survivehr_pretrain(
  events     = events,
  config     = cfg,
  event_vocab = my_vocab
)
```

This is useful when pooling data across sites: build a common vocabulary
from the union of all site codelists, then fix it so that model weights are
comparable.

---

## FastEHR column aliases

All functions accept the uppercase column names used by FastEHR:

| RSurvivEHR name | FastEHR alias |
|---|---|
| `patient_id` | `PATIENT_ID` |
| `event` | `EVENT` |
| `age` | `DAYS_SINCE_BIRTH` |
| `value` | `VALUE` |
| `target_event` | `TARGET_EVENT` / `EVENT` |
| `target_age` | `TARGET_AGE` / `DAYS_SINCE_BIRTH` |

When `DAYS_SINCE_BIRTH` is used, remember to set `time_scale = 1825.0` in
`survivehr_config()`.

---

## Full 6-patient worked example

This section walks through a complete analysis on 6 synthetic patients.

```{r}
library(RSurvivEHR)

# ── 1. Data ─────────────────────────────────────────────────────────────────
all_events <- data.frame(
  patient_id = c(1,1,1,1,1,1,  2,2,2,2,2,  3,3,3,3,3,  4,4,4,4,4,  5,5,5,5,5,5,  6,6,6,6,6,6),
  event = c(
    "HYPERTENSION","STATIN","T2D","BP_CHECK","HBA1C","METFORMIN",
    "T2D","METFORMIN","HBA1C","HYPERTENSION","STATIN",
    "STATIN","T2D","METFORMIN","BP_CHECK","HBA1C",
    "HYPERTENSION","AMLODIPINE","BP_CHECK","STATIN","T2D",
    "T2D","STATIN","HBA1C","HYPERTENSION","METFORMIN","CVD",
    "T2D","HYPERTENSION","STATIN","BMI","METFORMIN","DEATH"
  ),
  age = c(
    50.0, 50.5, 52.0, 52.1, 52.3, 52.5,
    45.0, 45.3, 46.0, 47.5, 48.0,
    58.0, 60.1, 60.4, 61.5, 62.0,
    55.0, 55.4, 56.0, 57.2, 58.0,
    48.0, 48.6, 49.0, 49.1, 50.5, 51.2,
    60.0, 60.8, 62.5, 62.8, 63.0, 64.1
  ),
  value = c(
    NA,  NA,  NA,  152, 54,  NA,   # p1: BP_CHECK=152 mmHg, HBA1C=54 mmol/mol
    NA,  NA,  61,  NA,  NA,        # p2: HBA1C=61 mmol/mol
    NA,  NA,  NA,  145, 58,        # p3: BP_CHECK=145 mmHg, HBA1C=58 mmol/mol
    NA,  NA,  168, NA,  NA,        # p4: BP_CHECK=168 mmHg (hypertensive)
    NA,  NA,  67,  NA,  NA,  NA,   # p5: HBA1C=67 mmol/mol (poorly controlled)
    NA,  NA,  NA,  32.1,NA,  NA    # p6: BMI=32.1 kg/m² (obese)
  )
)

static_covariates <- data.frame(
  patient_id     = 1:6,
  SEX            = c("M","F","M","F","M","F"),
  ETHNICITY      = c("White","Asian","White","White","Black","White"),
  SMOKING_STATUS = c("Never","Never","Ex","Never","Ex","Never"),
  IMD            = c(3L, 1L, 5L, 2L, 4L, 3L),
  YEAR_OF_BIRTH  = c(1965L,1972L,1955L,1960L,1968L,1958L)
)

# ── 2. Config ────────────────────────────────────────────────────────────────
cfg <- survivehr_config(
  block_size = 32, n_layer = 2, n_head = 2, n_embd = 64,
  epochs = 2, batch_size = 4,
  surv_layer = "competing-risk",
  time_scale = 1.0
)

# ── 3. Pre-train on all patients ─────────────────────────────────────────────
pt <- survivehr_pretrain(all_events, static_covariates, cfg)
pt$event_vocab
# <PAD>=0, <UNK>=1, AMLODIPINE=2, BMI=3, BP_CHECK=4, CVD=5,
# DEATH=6, HBA1C=7, HYPERTENSION=8, METFORMIN=9, STATIN=10, T2D=11

# ── 4. Fine-tune on CVD / DEATH (patients 5 & 6) ────────────────────────────
# Context = all events BEFORE the labelled outcome age (leakage-free)
ft_events <- all_events[!all_events$event %in% c("CVD","DEATH"), ]

targets <- data.frame(
  patient_id   = c(5L, 6L),
  target_event = c("CVD", "DEATH"),
  target_age   = c(51.2, 64.1)
)
# Patient 5: last context age = 50.5  → delta = (51.2 - 50.5) / 1.0 = 0.7
# Patient 6: last context age = 63.0  → delta = (64.1 - 63.0) / 1.0 = 1.1

ft <- survivehr_finetune(
  events            = ft_events,
  targets           = targets,
  outcomes          = c("CVD", "DEATH"),
  risk_model        = "competing-risk",
  static_covariates = static_covariates,
  config            = cfg,
  pretrained_model  = pt
)

# ── 5. Predict on context patients ───────────────────────────────────────────
preds <- survivehr_predict(ft, ft_events, static_covariates)
preds

# ── 6. Save / load round-trip ────────────────────────────────────────────────
tmp <- tempfile(fileext = ".pt")
survivehr_save_model(ft, tmp)
ft2 <- survivehr_load_model(tmp)
identical(ft2$event_vocab, ft$event_vocab)  # TRUE
unlink(tmp)
```

---

## Python environment setup

RSurvivEHR uses `reticulate` to call the vendored backend.  The first call
to any training function will attempt to import PyTorch, pandas, and numpy.
If these are not available in the current Python environment, install them:

```r
reticulate::py_install(
  c("torch", "pandas", "numpy", "lightning"),
  pip = TRUE
)
```

To pin a specific Python interpreter:

```r
reticulate::use_python("/path/to/python", required = TRUE)
```

See `vignette("python_dependencies", package = "reticulate")` for full
details.
