---
title: "Advanced topics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Token policy in depth

RSurvivEHR constructs the event vocabulary as:

```
Index  Token
─────  ─────────────────────────────────────────
  0    <PAD>   always present; padding_idx for the embedding layer
  1    <UNK>   when include_unk = TRUE (default)
  2    <CLS>   when include_cls_sep = TRUE
  3    <SEP>   when include_cls_sep = TRUE
  2+   clinical events, ordered by frequency (most common first)
```

`<PAD>` is always index 0 because SurvivEHR's `SplitDynamicEmbeddingLayer`
uses `padding_idx = 0` internally.

At prediction time any event code not present in `event_vocab` is replaced
by `<UNK>` (index 1), provided `include_unk = TRUE`.

> **Pre-train only.** `include_unk` and `include_cls_sep` are fixed when the
> vocabulary is built during `survivehr_pretrain()`.  The values stored in the
> bundle are reused automatically at fine-tune and prediction time; any
> `include_unk` / `include_cls_sep` values in a fine-tune config are ignored.

---

## `surv_layer` and `risk_model`

`surv_layer` (in `survivehr_config()`) and `risk_model` (an argument to
`survivehr_finetune()`) look similar but refer to **different survival heads**:

| Parameter | Where | Controls |
|---|---|---|
| `surv_layer` | `survivehr_config()` | The **backbone time-to-next-event head**.  Pre-training is not just "predict which event comes next" — it also trains a survival ODE to predict *when* the next event occurs, using age as the time axis.  `surv_layer` frames this as competing-risk (one event can be next; all others compete) or single-risk (each token modelled independently 1-vs-all).  Active during **both** pre-training and fine-tuning; the backbone architecture must remain consistent. |
| `risk_model` | `survivehr_finetune()` arg | The **outcome-level head** added on top of the backbone — predicts patient-level CIF curves for the specific labelled outcomes.  A brand-new, freshly initialised `nn.Module` that is never touched during pre-training and sits beside the backbone.  Entirely independent of `surv_layer`. |

Because they control completely separate modules, **`surv_layer` and
`risk_model` can be set differently without any conflict**.  For example,
you can pretrain with `surv_layer = "competing-risk"` and then fine-tune
with `risk_model = "single-risk"`, or vice versa.

In practice, leave `surv_layer = "competing-risk"` (the default) unless you
have a specific reason to change it, and use `risk_model` in
`survivehr_finetune()` to choose `"competing-risk"` (multiple outcomes) or
`"single-risk"` (one outcome).

---

## Age normalisation

All raw ages are divided by `time_scale` before entering the model:

$$a_{\text{norm}} = \frac{a_{\text{raw}}}{\texttt{time\_scale}}$$

**During pre-training**, this is the only role of `time_scale` — it controls
how ages are scaled before being fed to the transformer.  There is no
prediction window during pre-training.

**During fine-tuning**, `time_scale` additionally defines the prediction
window: the target time delta is normalised by the same divisor,

$$\delta_{\text{norm}} = \frac{a_{\text{target}} - a_{\text{last context}}}{\texttt{time\_scale}}$$

so the ODE integrates over $[0, 1]$ which maps to $[0, \texttt{time\_scale}]$
raw years (or whatever unit your ages are in).

Use `time_scale = 1.0` (default) when ages are in **years**.  
Use `time_scale = 365.25` when ages are in **days** (e.g. 18 262 days ≈ 50 years).

The `time_scale` is stored in the model bundle and automatically applied at
fine-tune and predict time, so you only need to set it once in
`survivehr_config()`.

---

## Static covariates encoding

Static features are encoded into a flat numeric vector per patient:

| Column type | Encoding |
|---|---|
| Character / factor / < 80 % numeric | One-hot via `pd.get_dummies` |
| Fully numeric (int or float) | Passed through as-is |
| Patient missing from `static_df` | All zeros |

> **Column naming rule**: `patient_id` (or the uppercase alias `PATIENT_ID`)
> is the only reserved name in the static table.  All other column names are
> **user-defined** — lowercase is recommended for consistency (e.g. `sex`,
> `imd`, `year_of_birth`).  The column names become the basis for encoded
> feature names after one-hot expansion (e.g. `sex` → `sex_F`, `sex_M`).
> This is in contrast to the **events** and **targets** tables, which require
> fixed lowercase column names (`patient_id`, `event`, `age`, `value`) —
> uppercase aliases are accepted for backward compatibility but lowercase is
> the canonical form.  See the *Column name aliases* section below.
> The only constraint for static columns is **consistency**: the exact same
> names must be used at pretrain, fine-tune, and predict time.

```{r}
static_covariates <- data.frame(
  patient_id     = 1:3,
  sex            = c("M",     "F",     "M"),       # → sex_F, sex_M
  ethnicity      = c("White", "Asian", "White"),   # → ethnicity_Asian, ethnicity_White
  smoking_status = c("Never", "Ex",    "Current"), # → smoking_status_Current, ..., Never
  imd            = c(3L, 1L, 5L),                 # numeric, passed through
  year_of_birth  = c(1965L, 1972L, 1955L)         # numeric, passed through
)
```

The resulting static embedding for patient 1 (`M`, `White`, `Never`, imd=3, year_of_birth=1965)
would be:

```
[sex_F=0, sex_M=1, ethnicity_Asian=0, ethnicity_White=1,
 smoking_status_Current=0, smoking_status_Ex=0, smoking_status_Never=1,
 imd=3, year_of_birth=1965]
```

The **same** encoding scheme (same column order, same dummy categories) is
derived at pre-train time and then **reused** at fine-tune and predict time
from the saved bundle.

---

## Using a fixed vocabulary

If you have pre-computed a vocabulary (e.g., from a previous run or from a
shared codelist), pass it as `event_vocab`:

```{r}
# Build vocab manually
my_vocab <- c("<PAD>" = 0L, "<UNK>" = 1L,
              "HYPERTENSION" = 2L, "METFORMIN" = 3L,
              "STATIN" = 4L, "T2D" = 5L)

pt_model <- survivehr_pretrain(
  events     = events,
  config     = cfg,
  event_vocab = my_vocab
)
```

This is useful when pooling data across sites: build a common vocabulary
from the union of all site codelists, then fix it so that model weights are
comparable.

---

## Column name aliases

All functions accept lowercase or uppercase column names:

| Standard name | Accepted alias |
|---|---|
| `patient_id` | `PATIENT_ID` |
| `event` | `EVENT` |
| `age` | `DAYS_SINCE_BIRTH` |
| `value` | `VALUE` |
| `target_event` | `TARGET_EVENT` / `EVENT` |
| `target_age` | `TARGET_AGE` / `DAYS_SINCE_BIRTH` |

The `DAYS_SINCE_BIRTH` alias exists for compatibility with datasets where ages
are stored as days rather than years.  If you use it, set
`time_scale = 365.25` in `survivehr_config()` so ages are normalised to
year-scale before entering the model.

---

## Full worked example

This section reproduces the complete pipeline from the Getting Started vignette
in a single self-contained block, demonstrating every best practice:
leakage-free context construction, targets defined before context filtering,
and the competing-risk fine-tune on two outcomes.

```{r}
library(RSurvivEHR)

# ── 1. Data (same population as the Getting Started vignette) ────────────────
events_pop <- data.frame(
  patient_id = c(rep(1,4), rep(2,6), rep(3,6), rep(4,4), rep(5,6),
                 rep(6,4), rep(7,4), rep(8,4), rep(9,6), rep(10,4)),
  event = c(
    "HYPERTENSION","STATIN","BP_CHECK","CVD",
    "HYPERTENSION","BP_CHECK","T2D","METFORMIN","HYPERTENSION","CVD",
    "HYPERTENSION","BP_CHECK","STATIN","T2D","BP_CHECK","CVD",
    "HYPERTENSION","STATIN","T2D","METFORMIN",
    "HYPERTENSION","BP_CHECK","T2D","HBA1C","METFORMIN","STATIN",
    "HYPERTENSION","AMLODIPINE","BP_CHECK","STATIN",
    "STATIN","T2D","HBA1C","METFORMIN",
    "HYPERTENSION","BP_CHECK","STATIN","T2D",
    "HYPERTENSION","BP_CHECK","T2D","METFORMIN","HBA1C","STATIN",
    "STATIN","BP_CHECK","HYPERTENSION","T2D"
  ),
  age = c(
    55.0, 55.5, 56.2, 58.0,
    44.0, 45.5, 48.0, 48.3, 50.5, 52.0,
    58.0, 59.5, 62.0, 63.5, 64.0, 65.5,
    45.0, 45.5, 47.0, 48.3,
    48.0, 49.0, 51.0, 52.0, 53.0, 54.5,
    60.0, 61.0, 62.3, 63.0,
    40.0, 42.0, 43.5, 44.8,
    58.0, 59.2, 60.0, 62.0,
    46.0, 47.5, 50.0, 52.0, 52.5, 54.0,
    44.0, 46.5, 47.0, 48.5
  ),
  value = c(
    NA,  NA,  148, NA,
    NA,  145, NA,  NA,  NA,  NA,
    NA,  158, NA,  NA,  162, NA,
    NA,  NA,  NA,  NA,
    NA,  152, NA,  68,  NA,  NA,
    NA,  NA,  155, NA,
    NA,  NA,  74,  NA,
    NA,  145, NA,  NA,
    NA,  138, NA,  NA,  71,  NA,
    NA,  138, NA,  NA
  )
)

static_pop <- data.frame(
  patient_id    = 1:10,
  SEX           = c("M","F","M","F","M","F","M","F","M","F"),
  ETHNICITY     = c("White","Asian","White","Black","White",
                    "Asian","White","White","Black","White"),
  IMD           = c(3L, 1L, 5L, 2L, 4L, 3L, 1L, 5L, 2L, 4L),
  YEAR_OF_BIRTH = c(1960L,1970L,1952L,1975L,1963L,1958L,1978L,1960L,1968L,1975L)
)

# ── 2. Config ─────────────────────────────────────────────────────────────────
# time_scale = 5.0 → 5-year prediction window; stored in pretrained bundle
# and inherited automatically at fine-tune and prediction time.
cfg <- survivehr_config(
  block_size = 64, n_layer = 2, n_head = 2, n_embd = 64,
  epochs = 10, batch_size = 4,
  surv_layer = "competing-risk",
  time_scale = 5.0
)

# ── 3. Pre-train on all 10 patients ───────────────────────────────────────────
pt <- survivehr_pretrain(events_pop, static_pop, cfg)
# Vocabulary ordered by frequency (most common first):
# <PAD>=0, <UNK>=1, HYPERTENSION=2, BP_CHECK=3, T2D=4, STATIN=5, ...
pt$event_vocab

survivehr_save_model(pt, "pretrain_backbone.pt")

# ── 4. Fine-tune: competing-risk CVD vs T2D (patients 1–6) ───────────────────
ft_static <- static_pop[static_pop$patient_id %in% 1:6, ]

# Define targets FIRST — the outcome age sets the context cut-off
targets <- data.frame(
  patient_id   = c(1L,    2L,    3L,    4L,    5L,    6L),
  target_event = c("CVD", "T2D", "T2D", "T2D", "T2D", "STATIN"),
  target_age   = c(58.0,  48.0,  63.5,  47.0,  51.0,  63.0)
)
# p1: CVD case.  p2–p5: T2D occurred first (CVD is the competing event for p2/p3).
# p6: censored — neither CVD nor T2D observed.

# Build context: remove both outcome codes AND events after the target age
ft_events <- events_pop[events_pop$patient_id %in% 1:6 &
                           !events_pop$event %in% c("CVD", "T2D"), ]
ft_events <- merge(ft_events, targets[, c("patient_id", "target_age")], by = "patient_id")
ft_events <- ft_events[ft_events$age < ft_events$target_age,
                        c("patient_id", "event", "age", "value")]

ft <- survivehr_finetune(
  events            = ft_events,
  targets           = targets,
  outcomes          = c("CVD", "T2D"),   # two competing outcomes
  risk_model        = "competing-risk",
  static_covariates = ft_static,
  config            = cfg,
  pretrained_model  = pt   # vocabulary, time_scale, and weights all inherited
)

# ── 5. Predict on the full cohort (outcome codes removed from context) ────────
pred_events <- events_pop[!events_pop$event %in% c("CVD", "T2D"), ]
preds <- survivehr_predict(ft, pred_events, static_pop)
# Columns: patient_id, CVD_cdf_last, CVD_auc, T2D_cdf_last, T2D_auc
preds

# ── 6. Save / load round-trip ─────────────────────────────────────────────────
tmp <- tempfile(fileext = ".pt")
survivehr_save_model(ft, tmp)
ft2 <- survivehr_load_model(tmp)
identical(ft2$event_vocab, ft$event_vocab)  # TRUE
unlink(tmp)
```

---

## Prediction outputs in depth

### Fine-tuned models — standard columns

`survivehr_predict()` on a fine-tuned bundle always returns one row per patient
with these columns per outcome:

| Column | Description |
|---|---|
| `{outcome}_cdf_last` | Cumulative incidence at `time_scale` (end of the prediction window). |
| `{outcome}_auc` | Area under the CDF integrated over `[0, time_scale]`. A single scalar risk score. |

### Fine-tuned models — time-specific risk (`eval_times`)

Pass a numeric vector of time points (in the same units as `age`) to get the
CIF at arbitrary horizons within the prediction window:

```r
# 1-, 2-, 3- and 5-year CVD / T2D risks (time_scale = 5.0)
preds_tp <- survivehr_predict(
  ft, pred_events, static_pop,
  eval_times = c(1, 2, 3, 5)
)
# Additional columns: CVD_cdf_t1, CVD_cdf_t2, CVD_cdf_t3, CVD_cdf_t5
#                     T2D_cdf_t1, T2D_cdf_t2, T2D_cdf_t3, T2D_cdf_t5
```

**Constraints**:

- Every value must be in `(0, time_scale]`. Requesting beyond `time_scale`
  extrapolates past the ODE integration grid and raises an error.
- Values are snapped to the nearest of the 1 000 internal grid points;
  resolution ≈ `time_scale / 999` (about 0.005 years for a 5-year window).

### Pre-train models — multi-step generation

`survivehr_predict()` on a pre-trained bundle autoregressively samples future
events.  `max_new_tokens` controls how many steps to generate; the result is
one row per patient per step:

```r
pt_preds <- survivehr_predict(pt, events_pop, static_pop, max_new_tokens = 3)
# Columns: patient_id, step, generated_token, generated_event,
#          generated_age, generated_value
```

| Column | Description |
|---|---|
| `step` | Generation step (1 = next predicted event, 2 = one after that, …). |
| `generated_event` | Decoded event name; `"<UNK>"` for out-of-vocabulary tokens. |
| `generated_age` | Predicted age in raw units (de-normalised: `age_norm × time_scale`). |
| `generated_value` | Predicted measurement value (see below); `NaN` for non-measurement events. |

### Pre-train models — predicting numeric measurement values

`survivehr_predict_value()` queries the Gaussian value regression head that is
trained alongside the survival backbone when `value_weight > 0`:

```r
# Predict blood-pressure reading at next BP_CHECK
bp_preds <- survivehr_predict_value(
  pt,
  events_pop[events_pop$event != "BP_CHECK", ],
  outcome_event     = "BP_CHECK",
  static_covariates = static_pop
)
# Columns: patient_id, outcome_event, predicted_value_mean, predicted_value_sd
```

The head returns `NaN` for events that never carried a non-`NA` `value` at
pre-training time (e.g. discrete diagnoses like `"CVD"`).

**Important — value scale**: the model applies no internal standardisation to
the `value` column.  The head is trained and predicts in whatever units you
supply.  For measurements on very different scales (e.g. BP in mmHg ~140 vs
HbA1c in mmol/mol ~70) or large absolute values, pre-normalise before training
and back-transform predictions afterwards:

```r
# Example: z-score per event type before training
scalers <- events_pop |>
  dplyr::filter(!is.na(value)) |>
  dplyr::group_by(event) |>
  dplyr::summarise(v_mean = mean(value), v_sd = sd(value))

events_scaled <- events_pop |>
  dplyr::left_join(scalers, by = "event") |>
  dplyr::mutate(value = dplyr::if_else(!is.na(value),
                                        (value - v_mean) / v_sd, NA_real_)) |>
  dplyr::select(patient_id, event, age, value)

# After survivehr_predict_value(), back-transform:
preds <- preds |>
  dplyr::left_join(scalers, by = c("outcome_event" = "event")) |>
  dplyr::mutate(
    predicted_value_mean = predicted_value_mean * v_sd + v_mean,
    predicted_value_sd   = predicted_value_sd   * v_sd
  )
```

Fine-tuned bundles retain the backbone's value regression head, so
`survivehr_predict_value()` works equally with pretrain and fine-tuned bundles.

---

## Python environment setup

RSurvivEHR uses `reticulate` to call the vendored backend.  The first call
to any training function will attempt to import PyTorch, pandas, and numpy.
If these are not available in the current Python environment, install them:

```r
reticulate::py_install(
  c("torch", "pandas", "numpy", "lightning"),
  pip = TRUE
)
```

To pin a specific Python interpreter:

```r
reticulate::use_python("/path/to/python", required = TRUE)
```

See `vignette("python_dependencies", package = "reticulate")` for full
details.
