---
title: "Getting started with RSurvivEHR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RSurvivEHR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE   # Python backend not available on CRAN check machines
)
```

**RSurvivEHR** trains transformer-based survival models on longitudinal EHR data
directly from R, with no Python experience required.  The full pipeline is:

```
events + static ──▶ survivehr_pretrain()  ──▶ pt_model
                                                  │
                                      survivehr_save_model()  ←── save for reuse
                                                  │
events + targets ─▶ survivehr_finetune()  ──▶ ft_model  (CR or SR)
                                                  │
new events ────────▶ survivehr_predict()  ──▶ predictions
```

---

## 1  Setup (one-time)

Install the package from GitHub:

```{r}
# install.packages("remotes")
remotes::install_github("PM-Cardoso/RSurvivEHR")
```

```{r}
# install.packages("pak")
pak::pkg_install("PM-Cardoso/RSurvivEHR")
```

Then set up the Python backend (only needed once):

```{r}
library(RSurvivEHR)

# Creates the "RSurvivEHR" Python virtualenv and installs all dependencies.
# Safe to call repeatedly — it is a no-op if the environment already exists.
survivehr_setup()
```

---

## 2  Input data

### Events data frame

One row per clinical event, sorted by age within each patient.

```{r}
events <- data.frame(
  patient_id = c(1, 1, 1, 1, 1,   2, 2, 2, 2, 2,   3, 3, 3, 3, 3),
  event      = c("HYPERTENSION", "STATIN",    "T2D",     "BP_CHECK",  "METFORMIN",
                 "T2D",          "METFORMIN", "HBA1C",   "HYPERTENSION","STATIN",
                 "STATIN",       "T2D",       "BP_CHECK","HBA1C",     "METFORMIN"),
  age        = c(50.0, 50.5, 52.0, 52.1, 52.3,
                 45.0, 45.3, 46.0, 47.5, 48.0,
                 58.0, 60.1, 61.5, 62.0, 62.3),
  value      = c(NA,   NA,   NA,   152,  NA,     # patient 1: BP_CHECK=152 mmHg
                 NA,   NA,   61,   NA,   NA,     # patient 2: HBA1C=61 mmol/mol
                 NA,   NA,   145,  58,   NA)     # patient 3: BP_CHECK=145 mmHg, HBA1C=58 mmol/mol
)
```

Required columns: `patient_id`, `event`, `age`.
FastEHR aliases are also accepted: `PATIENT_ID`, `EVENT`, `DAYS_SINCE_BIRTH`.

> Continuous readings (blood pressure, HbA1c, BMI, etc.) go in the `value`
> column alongside the event that recorded them.  Rows without a reading are
> `NA`.  Set `value_weight > 0` in `survivehr_config()` to train an auxiliary
> regression head on these values; leave at `0` (default) to use token
> sequences only.

### Static covariates data frame

One row per patient.  Categorical columns (< 80 % numeric) are one-hot encoded
automatically; numeric columns pass through unchanged.

> **Column naming**: `patient_id` is the only reserved name.  Every other
> column can be named freely — the backend encodes whatever it finds.  By
> contrast, the **events** and **targets** tables require fixed column names
> (`patient_id`, `event`, `age`, `value`; and `patient_id`, `target_event`,
> `target_age`, `target_value` respectively) or their FastEHR UPPER_CASE
> aliases.

```{r}
static_covariates <- data.frame(
  patient_id     = c(1, 2, 3),
  SEX            = c("M", "F", "M"),           # categorical → one-hot
  ETHNICITY      = c("White", "Asian", "White"),# categorical → one-hot
  IMD            = c(3L, 1L, 5L),              # numeric     → pass-through
  YEAR_OF_BIRTH  = c(1965L, 1972L, 1955L)      # numeric     → pass-through
)
```

> **Important**: always pass the **same columns in the same order** across
> pretrain, fine-tune, and prediction.  The model records the exact encoded
> column list at training time and raises a clear error if prediction data does
> not match.

### Validate your data

Run the validators before any training call to catch schema problems early —
wrong column names, unsorted ages, non-numeric values — with a clear R error
rather than a cryptic Python traceback.

```{r}
survivehr_validate_events(events)
survivehr_validate_static(static_covariates)
# [OK] Events: 15 rows, 3 patients. Columns present, ages numeric and time-ordered.
# [OK] Static covariates: 3 patients, 4 covariate column(s): SEX, ETHNICITY, IMD, YEAR_OF_BIRTH.
```

---

## 3  Configuration

`survivehr_config()` returns a named list consumed by all training functions.

```{r}
cfg <- survivehr_config(
  block_size    = 64,   # context window — sequences are padded/truncated to this
  n_layer       = 2,    # transformer blocks
  n_head        = 2,    # attention heads (n_embd must be divisible by n_head)
  n_embd        = 64,   # embedding / hidden dimension
  dropout       = 0,    # dropout rate (increase for regularisation)
  learning_rate = 3e-4,
  epochs        = 10,
  batch_size    = 4,
  surv_layer    = "competing-risk",  # or "single-risk" — see Section 5
  time_scale    = 1.0   # ages are in years; use 1825.0 for DAYS_SINCE_BIRTH
)
```

Key parameters:

| Parameter | Description |
|---|---|
| `block_size` | Context window length — sequences are padded/truncated to this. |
| `surv_layer` | `"competing-risk"` models multiple simultaneous outcomes; `"single-risk"` models one. |
| `time_scale` | Age divisor applied before the model. `1.0` for years, `1825.0` for days. Must match across pretrain/finetune/predict. |
| `include_unk` | If `TRUE` (default), a `<UNK>` token is reserved for unseen events at prediction time. |

---

## 4  Pre-training

Pre-training builds the vocabulary and trains the backbone transformer on
the full event history.

```{r}
pt_model <- survivehr_pretrain(
  events            = events,
  static_covariates = static_covariates,
  config            = cfg
)

# Vocabulary built alphabetically after reserved tokens:
# <PAD>=0  <UNK>=1  BP_CHECK=2  HBA1C=3  HYPERTENSION=4  METFORMIN=5  STATIN=6  T2D=7
pt_model$event_vocab
```

### Saving the pretrained model

Save the backbone now so you can load it later and fine-tune on different
outcome definitions without re-training from scratch.

```{r}
survivehr_save_model(pt_model, "pretrain_backbone.pt")

# Reload at any time — vocabulary, time_scale, and token policy are all preserved
pt_model <- survivehr_load_model("pretrain_backbone.pt")
```

---

## 5  Fine-tuning

Fine-tuning requires a `targets` data frame that labels the outcome for each
patient. Events in the fine-tuning set must **not** contain the outcome event
for labelled patients (leakage-free context).

Build a combined events/static table that includes the labelled patients:

```{r}
# Add two labelled patients: patient 4 developed T2D, patient 5 developed CVD
events_all <- rbind(
  events,
  data.frame(
    patient_id = c(4, 4, 4, 4, 4,   5, 5, 5, 5, 5, 5),
    event      = c("HYPERTENSION", "AMLODIPINE", "BP_CHECK", "STATIN", "T2D",
                   "T2D", "STATIN", "HBA1C", "HYPERTENSION", "METFORMIN", "CVD"),
    age        = c(55.0, 55.4, 56.0, 57.2, 58.0,
                   48.0, 48.6, 49.0, 49.1, 50.5, 51.2),
    value      = c(NA, NA, 168, NA, NA,        # p4: BP_CHECK=168 mmHg
                   NA, NA, 67,  NA, NA, NA)    # p5: HBA1C=67 mmol/mol
  )
)

static_all <- rbind(
  static_covariates,
  data.frame(
    patient_id    = c(4, 5),
    SEX           = c("F", "M"),
    ETHNICITY     = c("White", "White"),
    IMD           = c(2L, 4L),
    YEAR_OF_BIRTH = c(1960L, 1968L)
  )
)
```

Validate the extended dataset before fine-tuning:

```{r}
survivehr_validate_events(events_all)
survivehr_validate_static(static_all)
# [OK] Events: 26 rows, 5 patients. Columns present, ages numeric and time-ordered.
# [OK] Static covariates: 5 patients, 4 covariate column(s): SEX, ETHNICITY, IMD, YEAR_OF_BIRTH.
```

### 5a  Competing-risk fine-tuning

Use `risk_model = "competing-risk"` when two or more outcomes compete — i.e.
the occurrence of one prevents the other from being observed.  A separate ODE
survival head is fitted for each outcome.

```{r}
targets_cr <- data.frame(
  patient_id   = c(4L, 5L),
  target_event = c("T2D", "CVD"),
  target_age   = c(58.0, 51.2)
)
survivehr_validate_targets(targets_cr)

cfg_cr <- survivehr_config(
  block_size = 64, n_layer = 2, n_head = 2, n_embd = 64,
  dropout = 0, learning_rate = 3e-4, epochs = 10, batch_size = 4,
  surv_layer = "competing-risk", time_scale = 1.0
)

# Remove outcome event from context to prevent data leakage
ft_cr <- survivehr_finetune(
  events            = events_all[!events_all$event %in% c("CVD"), ],
  targets           = targets_cr,
  outcomes          = c("T2D", "CVD"),
  risk_model        = "competing-risk",
  static_covariates = static_all,
  config            = cfg_cr,
  pretrained_model  = pt_model   # vocabulary + weights inherited
)

cat("Fine-tune (CR) loss history:", unlist(ft_cr$history), "\n")
```

### 5b  Single-risk fine-tuning

Use `risk_model = "single-risk"` when there is a single endpoint of interest
and no competing events need to be modelled explicitly.

```{r}
targets_sr <- data.frame(
  patient_id   = c(4L, 5L),
  target_event = c("T2D", "T2D"),
  target_age   = c(58.0, 50.5)
)
survivehr_validate_targets(targets_sr)

cfg_sr <- survivehr_config(
  block_size = 64, n_layer = 2, n_head = 2, n_embd = 64,
  dropout = 0, learning_rate = 3e-4, epochs = 10, batch_size = 4,
  surv_layer = "single-risk", time_scale = 1.0
)

ft_sr <- survivehr_finetune(
  events            = events_all[events_all$event != "CVD", ],
  targets           = targets_sr,
  outcomes          = c("T2D"),
  risk_model        = "single-risk",
  static_covariates = static_all,
  config            = cfg_sr,
  pretrained_model  = pt_model
)

cat("Fine-tune (SR) loss history:", unlist(ft_sr$history), "\n")
```

| Setting | When to use |
|---|---|
| `"competing-risk"` | Multiple simultaneous endpoints (e.g. CVD and T2D compete). |
| `"single-risk"` | One endpoint, no competing events to account for. |

---

## 6  Save and load fine-tuned models

```{r}
survivehr_save_model(ft_cr, "model_cr.pt")
survivehr_save_model(ft_sr, "model_sr.pt")

# Reload — weights, vocabulary, static column schema, and time_scale are preserved
ft_cr2 <- survivehr_load_model("model_cr.pt")
ft_sr2 <- survivehr_load_model("model_sr.pt")
```

---

## 7  Prediction

```{r}
# Competing-risk predictions (T2D vs CVD)
preds_cr <- survivehr_predict(ft_cr2, events_all, static_all)
head(preds_cr)
# Columns: patient_id, T2D_cdf_last, T2D_auc, CVD_cdf_last, CVD_auc

# Single-risk predictions (T2D only)
preds_sr <- survivehr_predict(ft_sr2, events_all, static_all)
head(preds_sr)
# Columns: patient_id, T2D_cdf_last, T2D_auc
```

Output columns:

| Column | Description |
|---|---|
| `{outcome}_cdf_last` | Cumulative incidence at the last time step of the prediction window. |
| `{outcome}_auc` | Area under the CDF — a scalar summary of overall risk. |

### New patients at inference

Pass new patients with exactly the same static columns used at training time.
The model stores the exact encoded column list (including one-hot expanded
categories) and raises a clear error rather than silently producing wrong
predictions if there is a mismatch.

```{r}
new_patient <- data.frame(
  patient_id = c(99, 99, 99, 99),
  event      = c("T2D", "BP_CHECK", "METFORMIN", "ASPIRIN"),  # ASPIRIN → <UNK>
  age        = c(55.0,  55.3,       55.4,         56.1),
  value      = c(NA,    158,        NA,           NA)           # BP_CHECK=158 mmHg
)

# Must include the same columns as static_all: SEX, ETHNICITY, IMD, YEAR_OF_BIRTH
new_static <- data.frame(
  patient_id    = 99,
  SEX           = "M",
  ETHNICITY     = "White",
  IMD           = 2L,
  YEAR_OF_BIRTH = 1963L
)

pred_new <- survivehr_predict(ft_cr2, new_patient, new_static)
print(pred_new)
```

See the [Model architecture](model-architecture.html) article for the full
configuration reference and the official Gadd et al. 2025 hyperparameter table.

---

## References

Gadd, C. et al. (2025). *SurvivEHR: Transformer-based survival analysis on
electronic health records*. medRxiv.
<a href="https://doi.org/10.1101/2025.08.04.25332916" target="_blank" rel="noopener noreferrer">doi:10.1101/2025.08.04.25332916</a>
