---
title: "Evaluating predictions with Inter-Event Concordance (IEC)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating predictions with Inter-Event Concordance (IEC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE   # Python backend not available on CRAN check machines
)
```

After training a pretrain model with `survivehr_pretrain()`, you can evaluate how well
the model's risk scores rank competing next events using **Inter-Event Concordance (IEC)**.

IEC measures the quality of predicted risk rankings for next-event prediction in competing-risk settings.

---

## What is IEC?

Inter-Event Concordance (IEC) is a ranking-based metric that answers:
**How well does the model's predicted risk score rank the true next event among all possible events?**

For each transition in a patient's history:
1. Rank all possible next events by their predicted risk (low → high)
2. Find the rank position of the true observed event
3. Normalize to [0, 1]

**Interpretation:**
- IEC ≈ **1.0**: True event ranked among highest-risk events (excellent prediction)
- IEC ≈ **0.5**: True event ranked in middle of risk distribution (fair prediction)
- IEC ≈ **0.0**: True event ranked among lowest-risk events (poor prediction)

---

## Three evaluation functions

### 1. `survivehr_compute_iec()`: Pure metric calculation

Use this for **post-hoc analysis** of pre-computed risk scores:

```{r}
library(RSurvivEHR)

# Assume you have:
# - risk_matrix: shape (n_transitions, n_events)
# - observed_events: vector of 1-indexed event IDs (same length as nrow(risk_matrix))

# Basic IEC
result <- survivehr_compute_iec(
  risk_scores = risk_matrix,
  observed_events = observed_events
)
print(result)

# With stratification by event type
result <- survivehr_compute_iec(
  risk_scores = risk_matrix,
  observed_events = observed_events,
  stratify_by_event = TRUE,
  event_vocabulary = c("HTN", "DM", "MI", "CKD")
)
print(result)
```

**Output:**
- `mean_iec`: Overall IEC across all transitions
- `n_valid`: Number of successfully computed transitions
- `iec_values`: Per-transition IEC scores
- `by_event`: (if stratified) Per-event IEC breakdown

---

### 2. `survivehr_predict_event_risks()`: Extract risk scores

Use this **debug function** to inspect the model's raw risk score matrix:

```{r}
# Extract risk scores for a cohort
risks <- survivehr_predict_event_risks(
  model = pretrain_model,
  events = test_events,
  static = test_static,
  max_patients = NULL
)

# Output structure:
# risks$risk_matrix: shape (n_transitions, n_events)
# risks$patient_ids: IDs of patients included
# risks$event_vocab: Event vocabulary used by model
# risks$n_events: Total number of events
```

This is useful for:
- Debugging model predictions
- Custom downstream analysis
- Understanding which events the model scores highest/lowest

---

### 3. `survivehr_evaluate_iec()`: End-to-end evaluation

Use this for **production workflows**. It combines extraction + computation with batch processing:

```{r}
# End-to-end IEC evaluation with stratification
iec_eval <- survivehr_evaluate_iec(
  model = pretrain_model,
  events = test_events,
  static = test_static,
  batch_size = 32,
  stratify_by_event = TRUE,
  aggregate_only = TRUE
)

print(iec_eval)
```

**Features:**
- Automatically extracts risk scores in batches (memory efficient)
- Computes IEC per batch and aggregates
- Returns stratified results by event type
- Handles missing events gracefully

---

## Complete example workflow

```{r}
library(RSurvivEHR)

# 1. Setup and load data
survivehr_setup()

# Create example data (from data-pipeline vignette)
events_df <- # ... your event data
static_df <- # ... your static covariates

# 2. Train pretrain model
cfg <- survivehr_config(
  block_size = 128,
  n_layer = 4,
  n_head = 8,
  n_embd = 256,
  epochs = 10
)

pretrain_model <- survivehr_pretrain(
  events = events_df,
  static_covariates = static_df,
  config = cfg
)

# 3. Prepare evaluation set (next-event prediction dataset)
# Select transitions where we have ≥2 observed events per patient
validation_data <- events_df %>%
  arrange(patient_id, age) %>%
  group_by(patient_id) %>%
  mutate(
    event_num = row_number(),
    n_events = n()
  ) %>%
  ungroup() %>%
  filter(n_events >= 2)

# Context: all events except the last for each patient
test_context <- validation_data %>%
  group_by(patient_id) %>%
  filter(event_num < max(event_num)) %>%
  ungroup() %>%
  select(patient_id, event, age, value)

# Targets: the last event for each patient
test_targets <- validation_data %>%
  group_by(patient_id) %>%
  filter(event_num == max(event_num)) %>%
  ungroup() %>%
  select(patient_id, true_next_event = event)

# Static for evaluation set
test_static <- static_df %>%
  semi_join(test_targets, by = "patient_id")

# 4. Evaluate IEC
iec_result <- survivehr_evaluate_iec(
  model = pretrain_model,
  events = test_context,
  static = test_static,
  batch_size = 32,
  stratify_by_event = TRUE,
  aggregate_only = TRUE
)

# 5. Interpret results
cat("\n=== IEC Evaluation Results ===\n")
cat("Overall IEC:     ", sprintf("%.4f", iec_result$mean_iec), "\n")
cat("Valid transitions:", iec_result$n_valid, "/", iec_result$n_total, "\n")
cat("\nIEC by Event Type:\n")
print(iec_result$by_event)
```

---

## Interpreting stratified results

Stratification reveals if the model performs differently on common vs. rare events:

```{r}
# Example stratified output:
#     event mean_iec n_obs
#   1   HTN   0.725  1500    (common: lower IEC acceptable)
#   2    DM   0.680  1200    (common: slightly lower)
#   3    MI   0.821    45    (rare: higher IEC indicates good ranking)
#   4   CKD   0.605    32    (rare: lower IEC, harder to predict)
```

**Key insights:**
- If rare events have lower IEC, the model may struggle with limited training examples
- If common events have lower IEC, the model may be overconfident in common outcomes
- Disparities reveal potential for targeted fine-tuning or reweighting

---

## Best practices

1. **Always use a separate validation set** to avoid overfitting assessment
2. **Check both overall and stratified IEC** to identify event-specific patterns
3. **Compare across model configurations** (layers, heads, learning rate) to find the best architecture
4. **Use `aggregate_only=FALSE`** in `survivehr_evaluate_iec()` if you need per-transition scores for visualization
5. **Monitor IEC during fine-tuning** to track downstream task performance

---

## See also

- [Data pipeline](data-pipeline.html): Preparing input data
- [Getting started](getting-started.html): Model training workflow
- [Advanced topics](advanced-topics.html): Custom workflows and troubleshooting
