---
title: "Model architecture and parameter reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model architecture and parameter reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette is the complete reference for **RSurvivEHR**.  It explains what
the model does, how every `survivehr_config()` parameter affects training,
what values the original authors used in their large-scale UK primary-care
study (Gadd et al., 2026), and how to choose values for your own data.

**Jump to:**
[How the model works](#how-the-model-works) ·
[Quick-start configs](#quick-start-configurations) ·
[Upstream hyperparameters](#upstream-hyperparameters) ·
[Full config reference](#survivehr_config-full-parameter-reference) ·
[Architecture](#architecture-block_size-n_layer-n_head-n_embd) ·
[Regularisation](#regularisation-dropout) ·
[Optimisation](#optimisation-learning_rate-epochs-batch_size) ·
[Loss weights](#loss-weights-surv_layer-surv_weight-value_weight) ·
[Tokenisation](#tokenisation-include_unk-include_cls_sep) ·
[Time & age](#time--age-time_scale-and-outcome_horizon) ·
[Hardware](#hardware-device) ·
[Fine-tuning notes](#fine-tuning-notes)

---

## How the model works

SurvivEHR is a **GPT-style decoder-only transformer** pre-trained on sequences
of clinical events (diagnoses, prescriptions, measurements) via *causal
language modelling*.  After pre-training, the model encodes the temporal
structure of a patient's medical history into a dense representation.  A
**neural-ODE survival head** is then fine-tuned on labelled outcome data
(target event and age at event) to produce calibrated time-to-event curves.

```
EHR sequences --> Tokeniser --> Token + positional embedding (RoPE)
                                              |
                                  n_layer x NeoGPT block
                                  (MHA + SwiGLU MLP + Pre-LN)
                                              |
                                  Last attended hidden state
                                              |
                        +-----------------------------------------+
                        |  Competing-risk ODE head (one ODE/event)|
                        |  or                                      |
                        |  Single-risk ODE head (one ODE)         |
                        +-----------------------------------------+
                                              |
                                  CIF curves per time horizon
```

### Transformer block (NeoGPT)

RSurvivEHR always uses the **NeoGPT** block, which differs from the original
nanoGPT block in three ways:

- **Rotary Position Embeddings (RoPE)** — encodes relative token positions
  inside attention, without a separate positional embedding layer.
- **SwiGLU activation** — replaces GELU in the MLP for improved gradient flow.
- **Pre-LN** — LayerNorm is applied *before* each sub-layer (attention and MLP)
  rather than after, which stabilises training.

### Neural-ODE survival head

The survival head uses the **DeSurv** neural ODE framework.  For the
**competing-risk** setting, one ODE is instantiated per outcome event, each
with an encoder MLP, a lightweight ODE right-hand-side network, and a sigmoid
cumulative incidence function (CIF) output.  For **single-risk**, a single ODE
produces one survival curve.

The ODE integrates over a normalised time grid $[0, 1]$, which maps back to
$[0, \texttt{time\_scale}]$ in your original age units (see
[`time_scale`](#time--age-time_scale)).

### Vocabulary construction

The vocabulary is built automatically during `survivehr_pretrain()` and stored
inside every model bundle:

1. All unique `event` codes are collected from the events data frame.
2. Special tokens are prepended at fixed indices:
   - `<PAD>` → index 0 (always present; `padding_idx = 0`)
   - `<UNK>` → index 1 (when [`include_unk = TRUE`](#tokenisation-include_unk-include_cls_sep))
   - `<CLS>` → index 2 (when [`include_cls_sep = TRUE`](#tokenisation-include_unk-include_cls_sep))
   - `<SEP>` → index 3 (when `include_cls_sep = TRUE`)
3. Clinical event tokens are assigned indices ≥ 2 (or ≥ 4 with CLS/SEP),
   ordered by **frequency — most common event first** (lowest index).

At fine-tuning, outcome codes absent from the pre-training vocabulary are
**automatically appended** and the embedding layer is extended with a
weight-safe copy from the pre-trained backbone.

---

## Quick-start configurations

Copy the block that best matches your setup.  All parameters not shown use
their defaults (see [Full config reference](#survivehr_config-full-parameter-reference)).

### Laptop / small dataset (< 5 000 patients)

```{r}
library(RSurvivEHR)

cfg_small <- survivehr_config(
  block_size    = 64,
  n_layer       = 2,
  n_head        = 2,
  n_embd        = 64,
  dropout       = 0.1,    # regularise on small data
  learning_rate = 3e-4,
  epochs        = 20,
  batch_size    = 8,
  surv_layer    = "competing-risk",
  time_scale    = 1.0     # backbone normalisation (set outcome_horizon at fine-tune time)
)
```

### Workstation / medium dataset (5 000 – 50 000 patients)

```{r}
cfg_medium <- survivehr_config(
  block_size    = 128,
  n_layer       = 4,
  n_head        = 4,
  n_embd        = 256,
  dropout       = 0.0,
  learning_rate = 3e-4,
  epochs        = 10,
  batch_size    = 32,
  surv_layer    = "competing-risk",
  time_scale    = 1.0     # backbone normalisation (set outcome_horizon at fine-tune time)
)
```

### HPC / large dataset — upstream CPRD settings (> 100 000 patients)

```{r}
# Pre-training
cfg_pretrain <- survivehr_config(
  block_size    = 256,
  n_layer       = 6,
  n_head        = 6,
  n_embd        = 384,
  dropout       = 0.0,
  learning_rate = 3e-4,
  epochs        = 10,
  batch_size    = 64,
  surv_layer    = "competing-risk",
  time_scale    = 5.0
)

# Fine-tuning (larger context window, increased lr and batch size)
cfg_finetune <- survivehr_config(
  block_size    = 512,
  n_layer       = 6,
  n_head        = 6,
  n_embd        = 384,
  dropout       = 0.0,
  learning_rate = 5e-4,
  epochs        = 20,
  batch_size    = 512,
  surv_layer    = "competing-risk",
  outcome_horizon = 5.0  # 5-year prediction window; time_scale inherited from pretrained bundle
)
```

---

## Upstream hyperparameters

The table below shows the hyperparameters used in the original CPRD study
(Gadd et al., 2026) alongside the corresponding `survivehr_config()` parameter.
Upstream defaults are also available in the published configuration files
<a href="https://github.com/cwlgadd/SurvivEHR/blob/main/examples/modelling/SurvivEHR/confs/config_CompetingRisk11M.yaml" target="_blank" rel="noopener noreferrer"><code>config_CompetingRisk11M.yaml</code></a>
and
<a href="https://github.com/cwlgadd/SurvivEHR/blob/main/examples/modelling/SurvivEHR/confs/config_SingleRisk.yaml" target="_blank" rel="noopener noreferrer"><code>config_SingleRisk.yaml</code></a>.

| Component | Hyperparameter | `survivehr_config()` param | Pre-training | Fine-tuning |
|---|---|---|---|---|
| **Architecture** | Layers | [`n_layer`](#architecture-block_size-n_layer-n_head-n_embd) | 6 | 6 |
| | Attention heads | [`n_head`](#architecture-block_size-n_layer-n_head-n_embd) | 6 | 6 |
| | Hidden size | [`n_embd`](#architecture-block_size-n_layer-n_head-n_embd) | 384 | 384 |
| | Max sequence length | [`block_size`](#architecture-block_size-n_layer-n_head-n_embd) | 256 | 512 |
| **Regularisation** | Dropout | [`dropout`](#regularisation-dropout) | 0.0 | 0.0 |
| **Optimisation** | Batch size | [`batch_size`](#optimisation-learning_rate-epochs-batch_size) | 64 | 512 |
| | Epochs | [`epochs`](#optimisation-learning_rate-epochs-batch_size) | 10 | 20 |
| | Optimiser | AdamW (fixed) | AdamW | AdamW |
| | Learning rate | [`learning_rate`](#optimisation-learning_rate-epochs-batch_size) | 3 × 10^−4 | 5 × 10^−4 |
| **Loss weights** | Backbone survival head | [`surv_layer`](#loss-weights-surv_layer-surv_weight-value_weight) | `"competing-risk"` | `"competing-risk"` |
| | Survival loss weight | [`surv_weight`](#loss-weights-surv_layer-surv_weight-value_weight) | 1.0 | 1.0 |
| | Value regression weight | [`value_weight`](#loss-weights-surv_layer-surv_weight-value_weight) | 0.1 | — |
| **Tokenisation** | Reserve `<UNK>` token | [`include_unk`](#tokenisation-include_unk-include_cls_sep) | `TRUE` | (inherited) |
| | Add `<CLS>`/`<SEP>` tokens | [`include_cls_sep`](#tokenisation-include_unk-include_cls_sep) | `FALSE` | (inherited) |
| **Time / age** | Age normalisation divisor | [`time_scale`](#time--age-time_scale-and-outcome_horizon) | 1.0 | (inherited) |
| | ODE prediction window | [`outcome_horizon`](#time--age-time_scale-and-outcome_horizon) | — | 5.0 |
| **Hardware** | Compute device | [`device`](#hardware-device) | `"auto"` (GPU) | `"auto"` (GPU) |

> **Note on the optimiser:** the upstream CPRD study used separate backbone
> and head learning rates, LR schedulers, warmup steps, and early stopping.
> The RSurvivEHR backend uses a **single AdamW optimiser** over all parameters
> with a constant learning rate — no scheduler or early stopping is implemented.
> The `learning_rate` you set applies uniformly to the whole model.

---

## `survivehr_config()` — full parameter reference

The annotated call below shows every parameter with its RSurvivEHR default
and a note on the upstream value.

```{r}
library(RSurvivEHR)

cfg <- survivehr_config(
  # -- Transformer architecture -------------------------------------------
  block_size    = 128,       # context window (tokens); upstream pre-train: 256, fine-tune: 512
  n_layer       = 4,         # transformer blocks;      upstream: 6
  n_head        = 4,         # attention heads;         upstream: 6
  n_embd        = 256,       # embedding dimension;     upstream: 384

  # -- Regularisation ------------------------------------------------------
  dropout       = 0.0,       # applied to attention, MLP residuals; upstream: 0.0

  # -- Optimisation --------------------------------------------------------
  learning_rate = 3e-4,      # single AdamW LR for all parameters
  epochs        = 1,         # upstream pre-train: 10;  fine-tune: 20
  batch_size    = 16,        # upstream pre-train: 64;  fine-tune: 512

  # -- Loss weights --------------------------------------------------------
  surv_layer    = "competing-risk",  # backbone survival head; or "single-risk"
  surv_weight   = 1.0,       # weight on the survival loss term
  value_weight  = 0.0,       # weight on value regression loss (upstream pre-train: 0.1)

  # -- Tokenisation --------------------------------------------------------
  include_unk     = TRUE,    # reserve <UNK> for unseen events; upstream: TRUE
  include_cls_sep = FALSE,   # add <CLS>/<SEP> tokens; upstream: FALSE (GPT-style)

  # -- Age / time handling -------------------------------------------------
  time_scale      = 1.0,       # backbone age normalisation divisor (raw_age / time_scale)
  outcome_horizon = NULL,      # ODE prediction window (fine-tune only); NULL = inherit time_scale

  # -- Hardware ------------------------------------------------------------
  device        = "auto"     # "auto", "cpu", or "cuda"
)
```

---

## Parameter details

### `block_size` -- context window

| RSurvivEHR default | Upstream pre-training | Upstream fine-tuning | Notes |
|---|---|---|---|
| `128` | `256` | `512` | Fine-tuning uses a longer window to include more history. |

The **context window** is the maximum number of tokens (clinical events)
seen by the model at once.  Sequences longer than `block_size` are
*right-truncated* -- only the most recent events are retained.  Sequences
shorter than `block_size` are *left-padded* with `<PAD>` tokens (index 0).

> **Tip:** set `block_size` to the 95th percentile of sequence lengths in
> your dataset.

---

### `n_layer` -- number of transformer blocks

| RSurvivEHR default | Upstream (both stages) |
|---|---|
| `4` | `6` |

Each block consists of multi-head self-attention (MHA), a two-layer MLP,
LayerNorm (Pre-LN), and residual connections.  More layers gives greater
representational capacity but requires more data and compute.

---

### `n_head` -- number of attention heads

| RSurvivEHR default | Upstream (both stages) | Notes |
|---|---|---|
| `4` | `6` | Must divide `n_embd` evenly (384 / 6 = 64 per head). |

---

### `n_embd` -- embedding / hidden dimension

| RSurvivEHR default | Upstream (both stages) | Notes |
|---|---|---|
| `256` | `384` | Total params ~= 12 x n_layer x n_embd^2. |

Controls the width of every layer.  The MLP hidden dimension inside each
block is `4 x n_embd`.  Common sizes: `64` (debug), `128` (tiny), `256`
(small), `384` (medium, upstream), `512`-`768` (large).

---

### `dropout`

| RSurvivEHR default | Upstream |
|---|---|
| `0.0` | `0.0` |

Not needed when training on millions of patients.  Set `0.1`-`0.3` for
small datasets (< 10 000 patients).

---

### Optimisation: `learning_rate`, `epochs`, `batch_size`

#### `learning_rate`

| RSurvivEHR default | Upstream pre-training | Upstream fine-tuning |
|---|---|---|
| `3e-4` | 3 × 10^−4 | 5 × 10^−4 |

The optimiser is **AdamW**.  The `learning_rate` you supply is applied
uniformly to all model parameters via a single optimiser — there is no
separate backbone rate or learning rate schedule in the RSurvivEHR backend.
The upstream study used a lower backbone rate (5 × 10^−5) during fine-tuning
to preserve pre-trained weights; this is not currently implemented here.

#### `epochs`

| RSurvivEHR default | Upstream pre-training | Upstream fine-tuning |
|---|---|---|
| `1` | `10` | `20` |

Number of full passes over the training data.  Early stopping is not
implemented — set `epochs` explicitly.  Scale up proportionally for smaller
datasets where each epoch sees fewer gradient steps.

#### `batch_size`

| RSurvivEHR default | Upstream pre-training | Upstream fine-tuning |
|---|---|---|
| `16` | `64` | `512` |

The upstream fine-tuning batch of 512 requires multi-GPU hardware.  On a
single laptop GPU (8 GB VRAM), `batch_size = 16`–`32` with `block_size = 128`
is a safe starting point.

---

### Loss weights: `surv_layer`, `surv_weight`, `value_weight`

#### `surv_layer` — backbone time-to-next-event head

SurvivEHR pre-training goes beyond standard causal language modelling
(predicting *which* event is next).  The backbone also trains a survival
head to predict *when* the next event occurs, using age as the time axis.
`surv_layer` controls how that timing problem is framed:

| Value | Description |
|---|---|
| `"competing-risk"` (default) | Treats each sequence position as a genuine competing-risk problem: only one event can be next, so all other vocabulary events implicitly compete.  One neural ODE is instantiated per vocabulary token. |
| `"single-risk"` | Fits one independent ODE per vocabulary token in a 1-vs-all fashion, treating all other events as censored rather than competing. |

> **Important:** `surv_layer` controls the *backbone* head used in both
> pre-training and fine-tuning — the backbone architecture must remain
> consistent across both stages.  The *outcome-level* head that produces
> patient predictions is controlled separately by the `risk_model` argument
> to `survivehr_finetune()`.  Because they are entirely separate `nn.Module`
> objects, `surv_layer` and `risk_model` **can differ** — e.g. pre-train with
> `surv_layer = "competing-risk"` and fine-tune with
> `risk_model = "single-risk"` without conflict.

The upstream config uses the abbreviated codes `"cr"` and `"sr"` — both are
also accepted by RSurvivEHR.

#### `surv_weight` and `value_weight`

The total pre-training loss is a weighted sum of three terms:

| Loss term | Weight parameter | Default | Description |
|---|---|---|---|
| Causal LM | — (fixed at 1) | 1.0 | Predict the *identity* of the next event token. |
| Backbone survival | `surv_weight` | `1.0` | Predict *when* the next event occurs (neural ODE). |
| Value regression | `value_weight` | `0.0` | Predict the *numeric value* of measurement events (e.g. BMI). |

The upstream CPRD study used `surv_weight = 1.0` and `value_weight = 0.1` for
models that also learned measurement magnitudes.  Set `value_weight > 0` only
if your events data frame includes meaningful numeric `value` entries.

---

### Tokenisation: `include_unk`, `include_cls_sep`

> **Pre-training only.**  Both settings are fixed when the vocabulary is built
> during `survivehr_pretrain()` and stored in the model bundle.  Values in a
> fine-tune config are ignored.

#### `include_unk`

| RSurvivEHR default | Upstream |
|---|---|
| `TRUE` | `TRUE` |

Reserves `<UNK>` at index 1.  Any event code seen at fine-tune or prediction
time that was absent from the pre-training vocabulary is silently mapped to
`<UNK>`.  Always leave `TRUE` for production use.

#### `include_cls_sep`

| RSurvivEHR default | Upstream |
|---|---|
| `FALSE` | `FALSE` |

Prepends `<CLS>` and appends `<SEP>` around each patient sequence, following
the BERT / BEHRT convention.  Not used in the upstream GPT-style pre-training.
Set `TRUE` only if you are adapting RSurvivEHR for BERT-style models.

---

### Time & age: `time_scale` and `outcome_horizon`

These two parameters were previously a single value (`time_scale`) that
controlled both the backbone input scale and the prediction window.
They are now **independent**.

#### `time_scale` — backbone age normalisation

All raw event ages are divided by `time_scale` before entering the transformer:

$$\text{age\_norm} = \frac{\text{raw\_age}}{\text{time\_scale}}$$

This controls only how the backbone *sees* age values — it has nothing to do
with the prediction window.  **Must be identical** between pre-training and
fine-tuning.  The value is stored in the pretrained bundle and inherited
automatically — you do **not** need to set it in a fine-tune config.

| Age unit | Recommended `time_scale` |
|---|---|
| Years (e.g. `50.0`) | `1.0` — backbone sees year values directly |
| Days since birth (e.g. `18262`) | `365.25` — backbone sees normalised fractions |

#### `outcome_horizon` — ODE prediction window (fine-tune only)

The survival ODE integrates over a normalised grid $[0, 1]$, where $1.0$
maps back to `outcome_horizon` in your raw age units.  This is what
determines what `cdf_last` means and what values are valid in `eval_times`.

`outcome_horizon` can **differ freely from `time_scale`**.  The canonical
use case is:

```r
# Pre-train: year-by-year backbone normalisation
cfg_pt <- survivehr_config(time_scale = 1.0)

# Fine-tune: 5-year risk window, independent of time_scale
cfg_ft <- survivehr_config(outcome_horizon = 5.0)
```

| `outcome_horizon` | `cdf_last` meaning | Valid `eval_times` |
|---|---|---|
| `1.0` | 1-year risk | `(0, 1]` years |
| `5.0` | 5-year risk | `(0, 5]` years |
| `1826.25` | 5-year risk (ages in days) | `(0, 1826.25]` days |

If `outcome_horizon = NULL` (default), it inherits `time_scale` for backward
compatibility.  `outcome_horizon` is stored in every fine-tune bundle and used
automatically at prediction time — no need to specify it when calling
`survivehr_predict()`.

---

### Hardware: `device`

| Value | Behaviour |
|---|---|
| `"auto"` (default) | Uses GPU (CUDA) if available, otherwise falls back to CPU. |
| `"cpu"` | Forces CPU computation. |
| `"cuda"` | Forces GPU — raises an error if no CUDA device is available. |

---

## Fine-tuning notes

The same `survivehr_config()` object is passed to both `survivehr_pretrain()`
and `survivehr_finetune()`.  A few parameters behave differently at fine-tune
time:

| Parameter | Fine-tuning behaviour |
|---|---|
| `time_scale` | Inherited automatically from the pretrained bundle; value in fine-tune config is ignored. |
| `outcome_horizon` | **Fine-tune only.** Sets the ODE prediction window length independently of `time_scale`.  Stored in the fine-tune bundle; read automatically by `survivehr_predict()`.  Defaults to `time_scale` if not set. |
| `include_unk` | Inherited from the pretrained bundle vocabulary. |
| `include_cls_sep` | Inherited from the pretrained bundle vocabulary. |
| `block_size` | Can be increased at fine-tune time (e.g. 256 → 512) to provide a longer context. |
| `learning_rate` | Applied uniformly to all parameters. No separate backbone rate or scheduler is implemented. |
| `epochs` | Early stopping is not implemented; set this explicitly. |

The following Python-side options exist in the upstream backend but are **not
yet exposed through `survivehr_config()`** in RSurvivEHR:

| Option | Upstream default | Description |
|---|---|---|
| PEFT / adapter | Full fine-tune | `"fix"` freezes the backbone; `"adapter"` adds LoRA-style adapter layers. |
| Layer-wise LR decay | Disabled | Applies exponentially smaller learning rates toward the input layers. |

---

## References

Gadd, C., Gokhale, K., Acharya, A. et al. (2026). *SurvivEHR: a competing
risks, time-to-event foundation model for multiple long-term conditions from
primary care electronic health records.* npj Digital Medicine.
<a href="https://doi.org/10.1038/s41746-026-02709-z" target="_blank" rel="noopener noreferrer">doi:10.1038/s41746-026-02709-z</a>
